---
title: "GWAS JRL 2021"
author: "C.LE BOUAR ; C.MELOT ; L.RIZZI"
date: "20/06/021"
output: 
  html_document: default
editor_options: 
  chunk_output_type: console
---

# Context
Creation of the directory place 

```{r setup, echo=TRUE, message=FALSE, results='hide', cache=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache.lazy = FALSE)

#library("rstudioapi")   # mettre la directory en place
#setwd(dirname(getActiveDocumentContext()$path)) # Set working directory to source file location
#getwd()

```

```{r}
rm(list = ls())

library(anyLib)
anyLib(c("data.table", "apercu", "mlmm.gwas", "corpcor","plyr"))
library(stringr)

```

#Introduction
The number of mummies aphids (terminal, living or both) are studied and the link with one or several SNP is observed. 
##Model used
(mettre ici le paragraphe du mat & met)


##Genotype data
###SNP location on chromosomes

Markers located on the wheat genome are loaded. 


```{r }
# physical positions of SNP on the Zavitan WEW2 version
load("../data/GENOTYPES/BREEDWHEAT_on_durum_physic_WEW2.Rdata")
BLAST[1,]

```

Il y a 242382 SNP pour lesquels on a une position possible.

```{r }
dim(BLAST)
```


```{r }
# toutes les infos dispo sur les snp
load("../data/GENOTYPES/caract_SNP_ALL_PHYS.Rdata")

```

## Genotypic matrix
Lines was genotyped with a DNA chip on 420 000 SNP. But only a part of this data will be used. Usefull data are loaded here. 

```{r }
# breed wheat genotypes
file="../data/GENOTYPES/SG_EPO_complet.Rdata"
load(file)

dim(SG)
```
There are 476 lines on 168 725 SNP

Then only the SNP with an assumed physical position are kept.

```{r }
# optional Keep sNP only with an assumed physical position
liste<-which(colnames(SG) %in% BLAST[,1])
length(liste)
ap(liste)

SG<-SG[,liste]
dim(SG)

```


# Phenotypes 
## Data
BLUPS and BLUE for total mummies aphids are loaded here

```{r}
file<-"../data/PHENOTYPES/BlueBlupS2.csv"

myY  <-read.table(file, head = TRUE, sep=";", dec=",")
myY<-filter(myY,blup<2,8)
names(myY)[1]<-"Taxa"

dim(myY)
myY[1:10, 1]

# modification of the data because of a error of notation
myY[,1]<-str_replace(as.character(myY[,1]), "A", "4")

```


## Links between variable
```{r}
names(myY)
pairs(myY[,c(2,3)])
cor(myY[,c(2,3)])**2

```
BLUE and BLUPS are correlated.


# Keep Genotypes only if phenotyped
```{r}
SG<-SG[which(rownames(SG) %in% myY[,1]),]
ap(SG)
dim(SG)
```



```{r}

myY<-myY[which(myY[,1] %in% rownames(SG) ),]
dim(myY)
dim(SG)

```
The two number have to be equal

## Data pretreatment
```{r}
genot<-SG

class(genot)
dim(genot)
ap(genot)

```


### Imputation of Missing data 

- simple binomial Imputation

```{r na imp, cache = TRUE}

noms<-rownames(genot)
genot<-apply(genot,2,as.numeric)
rownames(genot)<-noms

genot.imp <- apply(genot, 2, function(x){
  freq <- table(x)
  x[is.na(x)] <- as.integer(names(which.max(freq)))
  return(x)
})

```


### Minimal allelic frequency


```{r freqs}
#Calcul of allelic frequency
p <- colMeans(genot.imp) / 2
q <- 1 - p
```


```{r maf, fig.height=4.5, fig.width = 5.5}
#Spatial representation of allelic frequency
maf <- apply(cbind(p, q), 1, min)
hist(maf, col = "grey", main = "", breaks = 50, xlim = c(0, 0.5))
```

Only allelic frequency superior to 5% are kept

```{r maf filter}
sum(maf < 0.5)
genot.ok <- genot.imp[, maf >= 0.05]
dim(genot.ok)
```
Only 65405 markers are kept.


### Physical map
Creation of the physical map
```{r import map}
map<-BLAST[which(BLAST[,1] %in% colnames(genot.ok)),c(1,2,6)]
names(map)<-c("SNP","Chr","Pos")
head(map)

```

Only selected SNP are kept and chromosome are called by number between 1 and 14.

```{r map filter}
map <- map[map$SNP %in% colnames(genot.ok), ]

map[,2] <-mapvalues(map[,2], from=c( "chrom", 
                          "chr1A", "chr1B", "chr2A", "chr2B",
                          "chr3A", "chr3B", "chr4A", "chr4B",
                          "chr5A", "chr5B", "chr6A", "chr6B",
                          "chr7A", "chr7B")
          , to=c("chrom", 1:14))

map[,2]<-as.numeric(map[,2])

dim(map)
head(map)
tail(map)
hist(map[,2],main="frequence chromosome")
```

SNP are classified according to their location on chromosome

```{r map sort}
map <- map[order(map$Pos), ]
map <- map[order(map$Chr), ]
head(map)
tail(map)

```

 
```{r clean genot}
#clean the memory
rm(genot, genot.imp, maf, p, q)

```


# Choice of the analysed variable

## Relation exploration
The number 3 is choosen to take the BLUPs

```{r}
names(myY)
#variable name i<-1
# BLUP i<-3
# BLUE i<-2
i<-3

```

## Creation of the phenotype vector
```{r import phenot}
hist(myY[,i], main = names(myY)[i])

y <- myY[,i]

# passage des noms dans le rowname de y
names(y) <- myY$Taxa

```

```{r phenot hist, fig.height=4.5, fig.width=5.5}

hist(y, col = "grey",  nclass=20, xlab = "BLUP Pucerons parasités", main=paste("Distribution of", names(myY)[i]))

summary(y)

```



# Data merge and viz

## Data check 

```{r plot match, fig.height=4.5, fig.width = 4.5}
# genotype are putting in the same order
genot.ok<-genot.ok[order(rownames(genot.ok)),]
y<-y[order(names(y))]

genot.loop<-genot.ok

# if they are missing data
if (length(which(is.na(y)))>0 ) { 
     liste<- which(is.na(y))
     y <- y[-liste] 
     genot.loop<-genot.loop[-liste,]
     #rownames(genot.loop)<-rownames(genot.ok)[-which(is.na(y))]
     }

dim(genot.loop)
```

The two files must be in the same order and must create a straight line. 

```{r plot match2, fig.height=4.5, fig.width = 4.5}
plot(match(rownames(genot.loop), names(y)))

# markers are putting in the same order in the map

genot.loop<-genot.loop[,map$SNP]
plot(match(map$SNP, colnames(genot.loop)))
```

## Kinship matric calcul

### K Van Raden

The apparent matrix of the model is calculated. The Van Raden method is used here. 

```{r , fig.height=4.5, fig.width = 4.5}
# apparent matrix calcul
p <- colMeans(genot.loop) / 2
q <- 1 - p

# The Van Raden matrix is centred
genot.scaled <- scale(genot.loop, center = 2 * p, scale = sqrt(2 * p * q))

K <- tcrossprod(genot.scaled) / ncol(genot.scaled)
K <- make.positive.definite(K)
```
The matrix gives the relation bitween two lines. 

```{r}
#dimension of the matrix
dim(K)
```

  
Here is the distribution of the values. 
Most lines have very low relatedness.

```{r , fig.height=4.5, fig.width = 4.5}
hist(K)
```

# Regression
The package mlmm.gwas. proposed by V.Segura and his co-workers is utilised. 



## package mlmm.gwas




```{r}
# from https://cran.r-project.org/web/packages/mlmm.gwas/vignettes/gwas-manual.html

# genotypes must be center 
genot.scaled <- scale(genot.loop, center = 2 * p) 

#In the function, y is the analysed phenotype ; K the Kinship matrix, genot.scaled the matrix with X ; threshold permit to controled the risk level; maxsteps gives the number of cofactor
mygwas.gwas <- mlmm_allmodels(y, list(genot.scaled), list(K), 
                               maxsteps = 4, threshold=1e-4)

res_mlmm <- mygwas.gwas
```

P-values are calculated for the 4 model and for all markers. 
A graph of -log10(Pvalues) for all markers is created. 
Two markers are above the threshold. 

```{r}
# manhattab plot without map
manhattan.plot(res_mlmm)
```

Then the best model is selected

```{r}
# model selection
sel_XX = frommlmm_toebic(list(genot.scaled), res_mlmm)
res_eBIC = eBIC_allmodels(y, sel_XX, list(K), ncol(genot.scaled))

res_eBIC
```
The best model is the one with the smallest eBIC. 
Here the model with the marker AX-89312681 has the smallest eBIC. It is significative. 
The same result is obtained with the analysis of BLUES



The threshold is fixed at 1e-5: 
```{r}
res_threshold <- threshold_allmodels(threshold=1e-5, res_mlmm)
```

The SNP AX-89312681 is the first too. 

```{r}
res_threshold

```


## Manhattan plot

Distribution of p-values on the genome are created.

```{r}
mip<-map
mip$Chr<-sprintf("%2d",mip$Chr)

manhattan.plot(res_mlmm, map = mip, steps = 2, hideCofactors = FALSE, chrToPlot = "all", unit = "bp")
```

There is a peak on the chromosome 1B, for the SNP AX-89312681


# Visualization of effects

Allelic effect (slope value) is calculated
```{r}
sel_XXclass = fromeBICtoEstimation(sel_XX, res_eBIC, res_threshold)

effects = Estimation_allmodels(y, sel_XXclass, list(K))
effects
```

At the locus AX-89312681, the genotype 00(aa) has a difference of 0.9232 mummies aphid compared to the genotype 11(AA). 


Informations on the SNP selected
```{r}
#information on the SNP
mark_list<-names(sel_XXclass)
mark_list[1]
```

Boxplot of the diffrence between the two homozygous is created
```{r}
# m is the numerous of the selected marker in the list
m<-1
genotypes.boxplot(genot.scaled, y , mark_list[m], effects)
```

The number of each class is calculated
```{r}
table(genot.loop[,mark_list[m]])

```

Allele with more mummies aphid is uncommon. 



```{r}
map[which(map$SNP == mark_list[m]),]

```

##  Analysis of close markers


```{r}
C<-cor(genot.ok[,mark_list[m]], genot.ok)**2

liste<-which(C>0.8)
liste<-colnames(C)[liste]

liste<-map[which(map$SNP %in% liste),]
liste
```


## est ce que les locus proches ont un effet ?
```{r}
genotypes.boxplot(genot.scaled, y , liste[2,1], effects)

```




